Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-08-27 write-up/covid-writeup.tex
Update covid-writeup.tex

There are two types of tests.  The first is

@@

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
$$
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
$$
Taking the expecation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparadness district).  Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

\subsubsection{An IPW estimator and statistical error decomposition.}
Given a selection propensity, define a weight~$W(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
$$
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\pi (x_i; \hat \theta)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k}{N \pi_k} (\bar y_k - FP),
$$
where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata and $d_k$ is the number of samples in strata $k$. Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

Let $\tilde I_j = I_j / (N \cdot \pi (X_j; \theta)) := I_j W_j$ for $j=1,\ldots,N$.  Then we can express $\bar y_n^\star$ as a weighted estimator $(1-FP-FN)^{-1} \sum_{i=1}^n $Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
$$
\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M} \times \frac{E_J[\tilde I_J]}{N}
$$
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.

@@

Until now, this paper focused primarily on selection bias in coronavirus case-count data from a purely statistical perspective.  Here, we extend the approach to whether it is possible bring in simple models from infectious diease epidemiology.  Our goal is simply an estimate of $\bar y_k$, the prevalence in strata $k$.

We rely heavily on recently available software~\cite{Song2020} for estimation of such models.

@@

\subsection{Doubly robust estimation}

$$
\sum_{k} P_k \hat y_k + \frac{1}{N} \sum_{i=1}^N I_i \frac{(Y_i - FP)/(1-FP-FN) - \hat y}{\pi_{K_i}}
$$
where $P_k$ is the population fraction in strata $k$. This estimator is ``doubly-robust'' in the sense that it is robust to either the SIR or the propensities being correctly specified.  Setting $\pi_i = n/N$ and $\hat y_k = 0$ yields the estimator $(\bar y_n - FP)/(1-FP-FN)$.

@@

{\bf where to describe difference from Chen}.

@@

>>>>

(Watch MICHAEL MINA VIDEO again)
There are two types of tests.  The first is

@@

When someone becomes infected, there will be an incubation period which may last between two to five days (cite: NPR) before the indivudal turn son the molecular/viral test.  Very quickly the viral load will be high.

Seven days post-symptom, the individual starts producing anti-bodies and so the seroprevalance test may be positive. A week later lasting a

@@

Here, we assume access to a probability sample from the same population on which the outcome and exact same set of covariates are measured.  Let $\tilde R_i$ be an indicator that the individual was included in the probability sample and $\tilde W_i$ be the probability of inclusion for the $i$th individual (i.e., equal to $n/N$ for a simple random sample of size $n$). The probability sample is used to construct a design-unbiased estimator of the second term
$$
\sum_{i=1}^N R_i \log \left( \frac{\pi (X_i; \theta)}{1-\pi(X_i; \theta)} \right)  + \sum_{i=1}^N \tilde R_i \tilde W_i \log ( 1 - \pi (X_i; \theta)).
$$
Taking the expectation of the second term with respect to the sampling design yields the second term in~\eqref{eq:propensity}.  Recall Indiana's probabilistic sample was a stratified random sampling design; therefore, the weights will depend on strata (i.e., 10 strata for each public health preparedness district).  Under simple random sampling and non-parametric selection propensity on a stratification variable, the weight is given by estimator of the $\hat \pi_k = (d_k / \tilde d_k) (n / N)$ where $\tilde d_k$ is the number of observations in strata $k$ in the random sample.

\subsubsection{An IPW estimator and statistical error decomposition.}
Given a selection propensity, define the inverse probability weight~$w(x) = \pi (x; \hat \theta)^{-1}$. Then the IPW estimator adjusted for measurement error is given by
$$
\bar y_n^\star
= \frac{1}{1-FP-FN} \cdot \frac{\sum_{i=1}^n w(x_i) (y_i - FP)}{\sum_{i=1}^n w (x_i)}
\stackrel{(2)}{=} \frac{1}{1-FP-FN} \sum_{k=1}^K \frac{d_k w_k}{w} (\bar y_k - FP),
$$
where the second equality (2) is under the assumption that $x_i$ is a stratification variable with $k$ indexing the strata, $w_k$ is the weight and $d_k$ is the number of samples in strata $k$, and $w = \sum_{k=1}^K d_k w_k$. Under simple random sampling with one strata~$\hat \pi_k = n/N$ and we recover the estimator $\tilde y_n$ from Section~\ref{section:imperfecttesting}.

Let $\tilde I_j = I_j  \cdot W(X_j)$ for $j=1,\ldots,N$.  Then the error when comparing weighted estimator~$\bar y_n^\star$ to the true prevalence $\bar Y$ can be expressed as:
$$
\rho_{\tilde I, Y} \times \sqrt{\frac{1-f+ CV^2_W}{f}} \times \sigma_{Y} \times \underbrace{\left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right] \times \frac{1}{1-(FP+FN)}}_{D_M}
$$
where $CV_W$ is the coefficient of variation (i.e., standard deviation/mean) of $W_J$ given $R_J = 1$.  See Appendix XX for the derivation.

@@

So far the focus has primarily been on selection bias in coronavirus case-count data from a purely survey sampling perspective.  Here, we consider  compartmental model approaches from infectious disease epidemiology.  Our primary objective is a model-based forecast of strata-level prevalence $\bar y_k$ using data up to but not including the current time.

Here, we present briefly a basic epidemiological state-space model -- the susceptible, infected, and removed (recovered and death) model, or SIR model.
The SIR model was originally proposed by (cite: Osthus, Forecasting seasonal influenza with a state-space sir model) with only one-dimensional time series of infected proportions. We refer the reader to Chapters 9-12 of [30] for a substantive introduction to state-space models.

More advanced versions of the SIR model can be used to make epidemiological forecasts.  Here, we focus on the SIER model due to availability of existing software.  We rely heavily on recently available software~\cite{Song2020} for estimation of such models.

The current prediction does not (A)

@@

\subsection{Doubly robust estimation.}
Two methods have been presented.  The first uses IPW to adjust for differential selection into the

Rather than relying on either independently, here we show how to combine these.
$$
\sum_{i=1}^N \mu (x_i; \hat \beta) + \frac{1}{\sum_{i=1}^n w(x_i)}
$$

There are again two options.  In the first,
$$
\sum_{k} P_k \hat y_k + \frac{1}{N} \sum_{i=1}^N I_i \frac{(Y_i - FP)/(1-FP-FN) - \hat y}{\pi_{K_i}}
$$
where $P_k$ is the population fraction in strata $k$. This estimator is ``doubly-robust'' in the sense that it is robust to either the SIR or the propensities being correctly specified.  Setting $\pi_i = n/N$ and $\hat y_k = 0$ yields the estimator $(\bar y_n - FP)/(1-FP-FN)$.

In the second, we rely on

@@

@@

{\bf where to describe difference from Chen. (1) weight normalized, (2) use aux info from multiple sources and use raking, (3) use a epi model rather than covariate pred model}.
